---
title: "Tennis_Proj"
output: html_document
date: "2026-02-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
source("elo_funcs.R")
library(dplyr)
```


```{r}
#load in data
data <- read.csv("match_data.csv")
new_dat <- read.csv("2026_aus_open.csv")

data$X <- NULL

data <- rbind(data, new_dat) #combine old with new updated data
tail(data, 500)
```

```{r}
data$tourney_date <- as.Date(as.character(data$tourney_date), format = "%Y%m%d") #changing date column to date format
#ensuring df is ordered by date. Should be already but just to make sure
data <- data[order(data$tourney_date, decreasing = FALSE), ]

```

```{r}
tail(data, 500)
```

```{r}
#standardizing player names
data$winner_name <- tolower(data$winner_name)
data$loser_name <- tolower(data$loser_name)

players <- unique(c(data$winner_name, data$loser_name)) #list of every player, winners and loser

players_elo <- data.frame(players = players, elo = rep(1500, length(players))) #give each player as starting elo of 1500
head(players_elo) 
```

Function to calculate ELO ratings

```{r}
k <- 32 #32 is generally the best k for tennis

data$winner_elo <- NA #empty column for winnner elo
data$loser_elo <- NA #empty column for loser elo

for (i in 1:nrow(data)){ #for each row in our date, each match
  winner <- data$winner_name[i] #get the winner
  loser <- data$loser_name[i] #get the loser
  
  winner_elo <- players_elo$elo[players_elo$players == winner] #find current elo rating for winner
  loser_elo <- players_elo$elo[players_elo$players == loser] #find current elo rating for loser
  
  data$winner_elo[i] <- winner_elo #add winner elo to df
  data$loser_elo[i] <- loser_elo #add loser elo to df
  
  #update elo calculation
  new_elo <- elo.calc(wins.A = 1, #1 means win
                      elo.A = winner_elo, #winner prematch elo
                      elo.B= loser_elo, #loser prematch elo
                      k=k)
  
  players_elo$elo[players_elo$players == winner] <- new_elo[1, 1] #updates winner elo with new elo
  players_elo$elo[players_elo$players == loser] <- new_elo[1, 2] #updates loser elo with new elo
  
  
  
}
```



```{r}
tail(data)
```



Surface ELO rankings function


```{r}
players <- unique(c(data$winner_name, data$loser_name)) #unique players

players_surf_elo <- data.frame(players = players, elo = rep(1500, length(players))) #new empty df of players elos
head(players_surf_elo)

empty_elo <- data.frame(players = players, elo = rep(1500, length(players))) #empty elo df for reference
```


Function for surface elos
```{r}
update_surface_elo_rankings <- function(data, players_surf_elo, k=32) { #takes in data, and starting surface elo

  #create empty columns
  data$winner_surf_elo <- NA 
  data$loser_surf_elo <- NA
  
  #for each match
  for (i in 1:nrow(data)){
    winner <- data$winner_name[i] #store winner name
    loser <- data$loser_name[i] #store loser name
    
    winner_elo <- players_surf_elo$elo[players_surf_elo$players == winner] #find current elo for winner
    loser_elo <- players_surf_elo$elo[players_surf_elo$players == loser] #find current elo for loser
    
    data$winner_surf_elo[i] <- winner_elo #pre match ratings
    data$loser_surf_elo[i] <- loser_elo
    
    #elo calculations using elos pre match
    new_elo <- elo.calc(wins.A = 1,
                        elo.A = winner_elo,
                        elo.B= loser_elo,
                        k=k)
    
    players_surf_elo$elo[players_surf_elo$players == winner] <- new_elo[1, 1] #update winner surface elo
    players_surf_elo$elo[players_surf_elo$players == loser] <- new_elo[1, 2] #update loser surface elo
     
    
    
  }
 
  
  return(list(match_results = data, final_rankings=players_surf_elo)) #returns df with match results and new surface ratings
}
```

Function to handle each surface
```{r}
surface_process <- function(surf_name, full_data, starting_elo, k=32) {#takes in surface name, match data, and starting elos
  
  sub_df <- full_data[full_data$surface == surf_name, ] #dataframe filtered for surface
  
  results <- update_surface_elo_rankings(sub_df, starting_elo, k=k) #update rankings using elo surface calc
  
  res <- results$match_results #store the match results
  
  #rename columns
  colnames(res)[colnames(res) == "winner_surf_elo"] <- "winner_surface_elo"
  colnames(res)[colnames(res) == "loser_surf_elo"] <- "loser_surface_elo"
  
  return(res)
}


data$orig_id <- 1:nrow(data) #adding an id column to keep track of match order

#perform elo surface calc for each surface
grass_results <- surface_process("Grass", data, empty_elo)
clay_results <- surface_process("Clay", data, empty_elo)
carpet_results <- surface_process("Carpet", data, empty_elo)
hard_results <- surface_process("Hard", data, empty_elo)

#recombine all of the data
data <- rbind(grass_results, clay_results, carpet_results, hard_results)


data <- data[order(data$orig_id), ] #order by the match id created earlier
data$orig_id <- NULL #remove orig_id
```


```{r}
#convert loser seed to dbl dtype
data$loser_seed <- as.numeric(data$loser_seed)
```



Converting to long format, removing win and loss data, two rows for each match, outcome column 0=loss, 1=win. Essentially creating two dfs, one for the winners, one for the losers, and combining them.
```{r}
#I know it looks shitty. Started out small but ended up adding many more columns
long_data <- bind_rows( #bind rows stacks the two dataframes vertically
  
  #winner data
  data %>% transmute(match_id = row_number(), #transmute creates new df with column names and values
                     tourney_name, 
                     tourney_level,
                     date = tourney_date,
                     best_of,
                     round,
                     player_id = winner_id, 
                     player = winner_name,
                     player_hand = winner_hand,
                     player_height = winner_ht,
                     player_age = winner_age,
                     player_rank = winner_rank,
                     player_elo = winner_elo,
                     player_surf_elo = winner_surface_elo,
                     player_rank_points = winner_rank_points,
                     player_seed = winner_seed,
                     opponent_id = loser_id,
                     opponent = loser_name,
                     opponent_hand = loser_hand,
                     opponent_height = loser_ht,
                     opponent_age = loser_age,
                     opponent_rank = loser_rank,
                     opponent_elo = loser_elo,
                     opponent_surf_elo = loser_surface_elo,
                     opponent_rank_points = loser_rank_points,
                     opponent_seed = loser_seed,
                     surface,
                     indoor,
                     minutes, 
                     player_ace = w_ace, 
                     player_df = w_df,
                     player_svpt = w_svpt,
                     player_first_srv_in = w_1stIn,
                     player_first_srv_won = w_1stWon,
                     player_second_serv_won = w_2ndWon,
                     player_serve_gms = w_SvGms,
                     player_bp_saved = w_bpSaved,
                     player_bp_faced = w_bpFaced,
                     opponent_ace = l_ace, 
                     opponent_df = l_df,
                     opponent_svpt = l_svpt,
                     opponent_first_srv_in = l_1stIn,
                     opponent_first_srv_won = l_1stWon,
                     opponent_second_serv_won = l_2ndWon,
                     opponent_serve_gms = l_SvGms,
                     opponent_bp_saved = l_bpSaved,
                     opponent_bp_faced = l_bpFaced,
                     outcome = 1,),
  #loser data
  data %>% transmute(match_id = row_number(),
                     tourney_name, 
                     tourney_level,
                     date = tourney_date, 
                     best_of,
                     round,
                     player_id = loser_id, 
                     player = loser_name,
                     player_hand = loser_hand,
                     player_height = loser_ht,
                     player_age = loser_age,
                     player_rank = loser_rank,
                     player_elo = loser_elo,
                     player_surf_elo = loser_surface_elo,
                     player_rank_points = loser_rank_points,
                     player_seed = loser_seed,
                     opponent_id = winner_id,
                     opponent = winner_name,
                     opponent_hand = winner_hand,
                     opponent_height = winner_ht,
                     opponent_age = winner_age,
                     opponent_rank = winner_rank,
                     opponent_elo = winner_elo,
                     opponent_surf_elo = winner_surface_elo,
                     opponent_rank_points = winner_rank_points,
                     opponent_seed = winner_seed,
                     surface,
                     indoor,
                     minutes, 
                     player_ace = l_ace, 
                     player_df = l_df,
                     player_svpt = l_svpt,
                     player_first_srv_in = l_1stIn,
                     player_first_srv_won = l_1stWon,
                     player_second_serv_won = l_2ndWon,
                     player_serve_gms = l_SvGms,
                     player_bp_saved = l_bpSaved,
                     player_bp_faced = l_bpFaced,
                     opponent_ace = w_ace, 
                     opponent_df = w_df,
                     opponent_svpt = w_svpt,
                     opponent_first_srv_in = w_1stIn,
                     opponent_first_srv_won = w_1stWon,
                     opponent_second_serv_won = w_2ndWon,
                     opponent_serve_gms = w_SvGms,
                     opponent_bp_saved = w_bpSaved,
                     opponent_bp_faced = w_bpFaced,
                     outcome = 0,)
) %>% arrange(date)

long_data
```


```{r}
names(long_data)
```


New simple difference features
```{r}
long_data <- long_data %>% mutate(
  #rank differences
  rank_diff = player_rank - opponent_rank,
  rank_diff_abs = abs(player_rank - opponent_rank),
  
  #age differences
  age_diff = player_age - opponent_age,
  age_diff_abs = abs(player_age - opponent_age),
  
  #height differences
  height_diff = player_height - opponent_height,
  height_diff_abs = abs(player_height - opponent_height),
  
  #elo differences
  elo_diff = player_elo - opponent_elo,
  elo_diff_abs = abs(player_elo - opponent_elo),
  
  #rank points differences
  rank_points_diff = player_rank_points - opponent_rank_points,
  rank_points_diff_abs = abs(player_rank_points - opponent_rank_points),
  
  #seed differences
  seed_diff = player_seed - opponent_seed,
  seed_abs = abs(player_seed - opponent_seed),
  
  
)
```

```{r}
names(long_data)
```
A little more cleaning
```{r}
long_data$round[long_data$round == 'R256'] <- 'R128'
long_data$round[long_data$round == 'Fs'] <- 'F'



unique(long_data$round)
tail(long_data)
```



```{r}
write.csv(long_data, "tennis_extnd_feats_v2_with_aussie_data.csv", row.names = FALSE)
```









